---
title: "STAT_563_Project_2_Q3"
author: "Cecilia Albert-Black"
date: "2025-11-03"
output: html_document
---

This project integrated large-sample theory, likelihood-based inference, and re-sampling techniques in modern computational statistics. In this Markdown file, I will:

-   Examine the logistic distribution and its importance in statistical modeling and machine learning

-   compute confidence intervals based on Fisher information

-   Explore bootstrapping and the distribution of correlation matrices

-   Implement slice sampling for correlation uncertainty

### Wald Confidence Interval and Percentile Bootstrap Confidence Interval for the Mean

Using the simulated data from Q2 part b, compute:

-   the Wald CI based on $$
    \mathrm{Var}(X) = \frac{\pi^2 s^2}{3}
    $$

-   the percentile bootstrap CI for the mean

Then, compare shapes and widths using the Matlab module provided.

#### a. Re-generate simulated data

```{r}
# Load necessary library
library(stats)

# Step 1: Generate random numbers from a logistic distribution
set.seed(123)  # For reproducibility
n <- 200
true_mu <- 0
true_s <- 1

# Logistic random variables
x <- rlogis(n, location = true_mu, scale = true_s)
```

#### b. Wald CI

```{r}
# Estimate mean and scale
mu_hat <- mean(x)
s_hat  <- sd(x) * sqrt(3)/pi   # approximate logistic scale from SD

# Variance of the sample mean using logistic formula
var_mu <- (pi^2 * s_hat^2) / (3 * n)  # Var(mean) = Var(X)/n
se_mu  <- sqrt(var_mu)

# 95% Wald CI
z <- qnorm(0.975)
ci_wald <- mu_hat + c(-1,1) * z * se_mu
ci_wald
```

#### c. Percentile bootstrap CI for the mean

```{r}
set.seed(123)
B <- 1000  # number of bootstrap samples
mu_boot <- replicate(B, {
  xb <- sample(x, replace = TRUE)
  mean(xb)
})

# 95% percentile bootstrap CI
ci_boot <- quantile(mu_boot, probs = c(0.025, 0.975))
ci_boot
```

#### d. Compare shapes and widths
